---
title: "Computing Relative Weights in FSA"
author: "Derek H. Ogle"
date: "`r Sys.Date()`"
format: 
  html: 
    toc: true
    toc-depth: 2
    reference-location: margin
    fig-cap-location: bottom
    tbl-cap-location: top
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

# Introduction
Condition is a measure of the physical condition of a population of fish based on the relative heaviness or plumpness of the fish as compared to a standard. One measure is called **relative weight** and is simply 100 times the observed weight of a fish divided by a standard weight derived from an accepted model for that species and the individual's observed length; i.e.,

$$ Wr_i = 100\frac{W_i}{Ws_i} $$ {#eq-Wri}

where $Wr_i$ is the relative weight, $W_i$ is the observed weight, and $Ws_i$ is the standard weight for the $i$th fish. Individual relative weights are summarized to provide an overall measure of condition for the population. Relative weight and condition are explained in a variety of sources, including [Ogle (2016)](https://fishr-core-team.github.io/fishR/pages/books.html#introductory-fisheries-analyses-with-r). The goal of this vignette is to demonstrate how to calculate relative weights using functions in`FSA`.

The following packages are used herein. Note that the `FSA` functions described here were modified after version 0.9.6 and are thus **specific to FSA >v0.9.6**.

```{r}
#| label: setup
#| results: hide
#| message: false
library(FSA)
library(dplyr)  # mutate, select, group_by, summarize
```


# Looking Up Standard Weight Equations
Equations for computing the standard weight (i.e., $Ws_i$) from an individual fish's observed length (i.e., $L_i$) have been derived for a number of freshwater game fish in the United States, as well as several non-game fish in the United States and some other fish from outside of the United States. The specifics of these equations have been collated into the `WSlit` data.frame^[Specifics [here](https://fishr-core-team.github.io/FSA/reference/WSlit.html).] distributed with `FSA` and are most easily accessed with `wsVal()`. For example, the specifics of the standard weight equation for Bluegill are retrieved below.

```{r}
wsVal("Bluegill")
```

The results returned from `wsVal()` are:

- `species`: species of fish asked for. If left blank then a list of all species for which a standard weight equation exists in `WSlit` will be shown.
- `group`: the sub-group for the species. Some species have separate standard weight equations for sub-groups (e.g,. `male` or `female`, or `lentic` and `lotic`). If `group` does not appear in the output then there is no sub-group for that species (as illustrated here). The sub-group can be chosen with `group=` in `wsVal()` as demonstrated later.^[Leave `group=` at the default `NULL` for species without sub-group equations.]
- `measure`: the length measure used (will generally be `TL` for total length, but could be `FL` for fork length, `BL` for body length, or `CL` for caudal length depending on the species).
- `units`: the units for which the equation was developed. The default is `metric` which has length in mm and weight in grams. However, `English` can also be used which has lengths in inches and weight in pounds. These are the only two "units" for which the equation can be returned; thus, if you recorded lengths and weights in different units you will need to adjust your data accordingly (or adjust the standard weights computed from the equations (see below) accordingly).
- `ref`: the quantile used when developing the standard weight equation. The 75th percentile is used for most species, but some species have alternatives (50th or 25th percentile). Alternatives for a species can be chosen with `ref=` in `wsVal()`.^[For species with only one possible `ref`, then `ref=` should be left at the default `NULL`.]
- `method`: the method used to develop the standard weight equation. This will usually be `RLP` (regression-line-percentile) or `EmP` (empirical percentile) but is `Other` only for Bluegill, as shown here. Some (very few) species have equations for both methods. In these instances, you will need to choose which to use with `method=` in `wsVal()`.^[For species where only one method is available then leave `method=` at the default `NULL`.]
- `min.TL`: the minimum length for which the standard weight equation is appropriate.^[The postfix after the period will be replaced with the `measure` type (here `TL` but for some species `FL`, `BL`, or `CL`).]
- `max.TL`: the maximum length for which the standard weight equation is appropriate. Maximum lengths have not been specified for all species and, thus, may not appear in all outputs (e.g., for Bluegill here).
- `int`: the intercept, **on the common logarithm (i.e., log10) scale**, for the standard weight equation.
- `slope`: the slope, **on the log10 scale**, for the standard weight equation.
- `quad`: the coefficient for the quadratic term, **on the log10 scale**, of the standard weight equation. Most species will not have a quadratic term and, thus, this may not appear in many outputs.
- `source`: the literature source for the standard weight equation.
- `comments`: comments about the standard weight equation. Not all species have comments, thus, this may not appear in many outputs.

A simpler output of only the equation coefficients and the lengths for which it applies may be returned by including `simplify=TRUE` in `wsVal()`.^[I don't often use this, but it is available for a "cleaner" look.]

```{r}
wsVal("Bluegill",simplify=TRUE)
```

Blue Sucker is an example where some of the "optional" values are returned (e.g., `max.TL` and `quad`). The example below further illustrates that the results can be returned in "English" units for many species.

```{r}
wsVal("Blue Sucker",units="English")
```

Use of `wsVal()` requires spelling (and capitalizing) the name as it appears in `WSlit`. One can see all species names available in `WSlit` with `wsVal()` without any arguments.

```{r}
wsVal()
```

All parts of the species names in `WSlit` are capitalized (e.g., "Blue Sucker" and not "blue sucker" or "Blue sucker"). `wsVal()` will return an informative error message if your capitalization is not correct but the message will be less informative if your spelling is off.

```{r}
#| error: true
wsVal("Blue sucker")
wsVal("Blue suckr")
```

It is also not always obvious whether a species has equations for sub-groups or not. One way to deal with this is to just ask for the equation for your species of interest without using `group=`. If sub-groups exist then you will get an error message asking you to choose which sub-group to use.

```{r}
#| error: true
wsVal("Brown Trout")
```

Then try again by choosing the sub-group with `group=`.

```{r}
wsVal("Brown Trout",group="lentic")
```

These same species and sub-group combinations can be also accessed by combining the species name and sub-group name (in parenthesis) into the first argument (and then not using `group=`).

```{r}
wsVal("Brown Trout (lotic)")
```

The same general process can be used for species that have equations developed from multiple methods; though, only the sub-group identifiers can be combined (with parentheses) in the species name and used in the first argument.

```{r}
#| error: true
wsVal("Arctic Grayling")
wsVal("Arctic Grayling",method="EmP")
```

Or for multiple reference groups.

```{r}
#| error: true
wsVal("Ruffe")
wsVal("Ruffe",ref=50)
```

There are two species (Chinook Salmon and Striped Bass) that appear to have sub-groups (i.e., "landlocked") but only that one specific sub-group appears in `WSlit`. The standard weight literature for these species did not identify the equation as only being for "landlocked" populations. However, these entries were added to facilitate use when calculating proportional size distribution (PSD) metrics,^[See [this vignette]().] which were defined just for "landlocked" populations, and relative weight metrics with the same data.frame.^[As illustrated in [this companion vignette().] The standard weight equations for these species can be accessed by either just the species name or the species name with the group in parentheses. However, note that they are the exact same information.

```{r}
wsVal("Striped Bass")
wsVal("Striped Bass (landlocked)")
```

There are also a few species where an original standard weight equation has been revised in  the literature. The original and revised equations are available in `WSlit` with the revised equations accessed by just using the species name and the original equations accessed by appending "(original)" to the species name.

```{r}
wsVal("Spotted Bass")             # revised definitions
wsVal("Spotted Bass (original)")
```

We strongly urge you to have a good understanding of the standard weight literature for your species' of interest and make sure that `wsVal()` is returning the values that you expect.

# Calculate Individual Relative Weight
## For Typical Linear Equation
The specifics of the standard weight equation returned by `wsVal()` can be used to compute the standard weight for a fish given its observed length. As an example, suppose that the relative weight of a Largemouth Bass with an observed length of 350 mm and weight of 650 g is desired.

Begin by assigning the specifics of the standard weight equation for Largemouth Bass returned by `wsVal()` to an object (e.g., `wsLMB` here).

```{r}
( wsLMB <- wsVal("Largemouth Bass") )
```

The intercept and slope for the standard weight equation on the **log10-log10** scale can be extracted from this object.^[The double-brackets (i.e., `[[]]`) are used to remove the name (e.g., `int`) from the returned value.]

```{r}
wsLMB[["int"]]
wsLMB[["slope"]]
```

The standard weight is then computed from these results and the **log10-transformed** observed length as follows (with the result saved to an object, called `ex1` here).  Make sure to note the use of `10^` to back-transform the **log10-transformed** standard weight to a standard weight on the raw scale.

```{r}
( ex1 <- 10^(wsLMB[["int"]]+wsLMB[["slope"]]*log10(350)) )
```

This calculation suggests that the standard weight for a 350 mm Largemouth Bass is `r formatC(ex1,format="f",digits=1)` g. With this and @eq-Wri, the relative weight for this individual is computed (recalling that the observed weight was 650 g).

```{r}
100*650/ex1
```

This indicates that the individual is (slightly) heavier than a standard fish of the same length as this values is greater than 100.

## For Quadratic Equation
A similar process can be followed for a species where the standard weight equation includes a quadratic term. The only "trick" here is to include the quadratic term multiplied by the **square** of the log10-transformed observed length when computing the standard weight. This calculation is illustrated below with a 500 mm and 1010 g Blue Sucker.

```{r}
( wsBS <- wsVal("Blue Sucker",simplify=TRUE) )
( ex2 <- 10^(wsBS[["int"]]+wsBS[["slope"]]*log10(500)+wsBS[["quad"]]*log10(500)^2) )
100*1010/ex2
```

## Cautions
A few things to consider when calculating relative weights for individuals.

* Make sure the individual is longer than the minimum and shorter than the maximum (if given) length returned by `wsVal()`.
* Make sure to log10-transform the observed length when computing the standard weight. Note that all standard weight equations use log10 so never use the natural log with standard weight calculations.
* Make sure to back-transform (i.e., raise to the power of 10) the value so that the standard weight is on the raw rather than log10-transformed scale.
* Relative weights have traditionally been reported on a scale where the fraction is multiplied by 100 (i.e., so 100 means that the observed weight equals the standard weight). Thus, don't forget to multiply by 100 when computing the relative weight.

# Calculate Relative Weights for All of One Species
It is not common to compute the standard and relative weights for a single individual as was done in the previous section. Rather, it is more useful to calculate these values for all individuals in a sample, and then summarize those values for an overall assessment of the condition of those individuals.

Consider the `CiscoTL` data.frame distributed with the `FSAdata` package that contains the lengths (mm) and weights (g) of Cisco (*Coregonus artedii*) sampled from Trout Lake, WI, USA over a 25 year period. Note that there are lots of missing weights in this data.frame.

```{r}
data("CiscoTL",package="FSAdata")
peek(CiscoTL,n=10)
```

## "Manual" Calculations
The relative weight calculation begins by finding the specifics of the standard weight equation for Cisco.

```{r}
( wsC <- wsVal("Cisco") )
```

Two new variables -- `Ws` for standard weight and `Wr` for relative weight -- are added to the data.frame in three steps below. First, `Ws` is calculated using the coefficients from the standard weight equation and the log-transformed length variable as shown above (but within `mutate()` from `dplyr`). Second, if the observed length is less than the minimum length for which the standard weight equation should be applied, then the previously calculated `Ws` is replaced with an `NA` (for missing value).^[These two steps could have been combined by replacing `Ws` with the calculation in `ifelse()`.] Finally, the relative weight is computed from the observed weight and standard weight variables using @eq-Wri.

```{r}
CiscoTL <- CiscoTL |>
  mutate(Ws=10^(wsC[["int"]]+wsC[["slope"]]*log10(length)),
         Ws=ifelse(length<wsC$min.TL,NA,Ws),
         Wr=100*weight/Ws)
peek(CiscoTL,n=10)
```

## Using the `WrAdd()` Convenience Function
`wrAdd()` can be used to add a relative weight variable to a data.frame for **all** species in the data.frame for which a standard weight equation exists.^[Though its use is illustrated here on a data.frame with only one species. See the next sections for an example with more species.] The main argument to `wrAdd()` is a formula of the form `weight~length+species` where `weight` is the observed weight variable, `length` is the observed length variable, and `species` is the name of the species variable. One constraint here is that the names in `species` must be spelled (and capitalized) exactly as in `WSlit`. In these data, the species name is in `spname` but is in all capital letters as "CISCO", rather than the required "Cisco". `capFirst()`^[From `FSA`.] may be used to convert a word to a form where just the first letter is capitalized, and  is used below.^[Note that the `CiscoTL` data.frame was re-read so that the additions in the previous section were removed.]

```{r}
data("CiscoTL",package="FSAdata")
CiscoTL <- CiscoTL |>
  mutate(Wr=wrAdd(weight~length+capFirst(spname)))
peek(CiscoTL,n=10)
```

A more general approach to deal with species that are spelled differently than expected in `WSlit` is to provide a named list or vector that defines how the original species names (i.e., the values in the vector) relate to the species names required by `WSlit` (i.e., the names in the vector) to `thesaurus=`. `wrAdd()` will match the two names appropriately while creating the relative weight variable.

```{r}
CiscoTL <- CiscoTL |>
  mutate(Wr2=wrAdd(weight~length+spname,thesaurus=c("Cisco"="CISCO")))
peek(CiscoTL,n=10)
```

`thesaurus=` can be used even if only some of the species names are non-"standard." Additionally, the named list/vector in `thesaurus=` can contain names that don't exist in the original data.frame. Thus, a global thesaurus containing all species that *could* be encountered could be created, for example as an agency-wide definition, and used with a variety of specific data.frames. 

`wrAdd()` handles species that have more than one standard weight equation^[As described previously.] with `WsOpts=`. For example, consider adding a relative weight variable to the `RuffeSLRH92` data.frame distributed with `FSAdata`. Note that there is no species variable, which is needed by `wrAdd()`.^[Note that `wrAdd()` is more convenient for data.frames with multiple species as will be illustrated in the next section.] Additionally, for simplicity of presentation, I removed several other variables that are not needed for this example.

```{r}
data("RuffeSLRH92",package="FSAdata")
RuffeSLRH92 <- RuffeSLRH92 |>
  mutate(species="Ruffe") |>
  select(species,length,weight,sex)
peek(RuffeSLRH92,n=10)
```

Code similar to that used above for Cisco returns an error because there are standard weight equations for Ruffe for both 50th and 75th percentile references (see snippet below with the error message).

```{r}
#| error: true
RuffeSLRH92 <- RuffeSLRH92 |>
  mutate(Wr=wrAdd(weight~length+species))
```

To continue, `wrAdd()` must be instructed as to which of these two equations should be used to calculate the standard weight and, thus, relative weight. This is handled with a list in `WsOpts=` which will include the species name set equal to a list which identifies the specific equation to use (i.e., `ref=50` in this example).

```{r}
RuffeSLRH92 <- RuffeSLRH92 |>
  mutate(Wr=wrAdd(weight~length+species,WsOpts=list(Ruffe=list(ref=50))))
peek(RuffeSLRH92,n=10)
```

# Calculate Relative Weights for All of Multiple Species
The real value of `wrAdd()` is that it can be used to efficiently add a relative weight variable for multiple species in a single data.frame. This is illustrated below with examples for a variety of scenarios.

## "Good" Names and No Groups
`InchLake2` distributed with `FSAdata` contains lengths and weights for several species captured from Inch Lake. These data provide a simple example for using `wrAdd()` because all species names are spelled as required and none of the species have sub-groups.^[See the next section for how to deal with these issues.]

```{r}
data("InchLake2",package="FSAdata")
peek(InchLake2,n=10)
```

A complication here is that length and weight are in "mixed" units; i.e., inches for length and grams for weight. One or the other must be converted to the same unit type. Below length is converted to mm.

```{r}
InchLake2 <- InchLake2 |>
  mutate(length=length*25.4,
         Wr=wrAdd(weight~length+species))
peek(InchLake2,n=10)
```

Note that species without known standard weight equations in `WSlit` (e.g., Bluntnose Minnow) will have `NA` for the relative weight variable, as will individuals with lengths outside the range of lengths for which the standard weight equation should be applied.

## "Bad" Names and No Groups
Now consider the `BGHRfish` data.frame (distributed with the `FSAdata` package) that has the lengths (mm) and weights (g) of three species -- Smallmouth Bass, Largemouth Bass, and Bluegill --  from Big Hill Reservoir, KS. These four species do not have sub-groups defined in `WSlit`. However, observing the data below^[And [the documentation](https://fishr-core-team.github.io/FSAdata/reference/Herman.html).] shows that the species variable (`specCode`) contains codes for the species names rather than the names required by `PSDlit`.

```{r}
data(BGHRfish,package="FSAdata")
peek(BGHRfish,n=10)
```

One complicating factor with these data is that `wrAdd()` expects a string in `species` rather than a numeric like `specCode`. Thus, below `specCode` will be converted to a character before using `wrAdd()`.

One way to deal with the issue of "bad" species names is to use a named list or vector that defines how the names from `WSlit` should be matched to the names in the data.frame. For this purpose, the species names in `WSlit` are the names in the vector (i.e., before the `=`) and the species names in the data.frame are the values in the vector (i.e., after the `=`).^[The numeric species codes are in parentheses as `specCode` will be converted to a string before using `wrAdd()`.]

```{r}
thes <- c("Smallmouth Bass"="116","Largemouth Bass"="118","Bluegill"="122")
```

This list/vector is then given to `thesaurus=` in `wrAdd()` which will perform the name matching while creating the GH length categories.

```{r}
BGHRfish <- BGHRfish |>
  mutate(specCode=as.character(specCode),
         Wr=wrAdd(weight~length+specCode,thesaurus=thes))
peek(BGHRfish,n=10)
```

## "Bad" Names, Groups, and Multiple Equations
The use of `wrAdd()` can become complicated for data.frames with species names other than what `WSlit` expects, with species for which standard weight equations exist for sub-groups, especially if more than one sub-group is in the data, and for species with multiple standard weight equations. The hypothetical data set `PSDWRtest` distributed with `FSA` can be used to illustrate how to handle these "issues".

```{r}
peek(PSDWRtest,n=20)
```

`wrAdd()` will produce some informative error messages, but it is best that you have a full understanding of the issues that may arise with your data by carefully examining your data and understanding the standard weight (and Gabelhouse length categories) for the species in your data. The "issues" that need to be addressed with the `PSDWRtest` data are as follows:
  
  - "Bluegill Sunfish" was used rather than "Bluegill".
  - "Lean Lake Trout" was used rather than "Lake Trout".
  - Brook Trout sampled from a lotic ("Trout Lake") system, for which there are standard weight equations for sub-groups.
  - Brown Trout sampled from a lotic ("Trout Lake") and lentic ("Brush Creek") system, for which there are standard weight equations for sub-groups.
  - Muskellunge for which sex is known for some and not for others, and there is a desire to use the sex-specific standard weight equation when sex is known (and use the overall equation when it is not).
  - Walleye of sizes for which separate standard weight equations may be used.
  - Ruffe for which the reference quantile of the standard weight equation must be specified.

The easiest way to deal with most of these "issues" is to create a new "species" variable (i.e., `species2` below) that appends the specific groups in parentheses to the species name. There are a variety of ways to do this and which way (is best or works) may depend on the specifics of the situation. Here, we primarily use `case_when()` from `dplyr` with a series of statements that begin with a "condition" to the left of the `~` and new species "name" for that condition to the right of the `~`. The `.default=species` at the end puts the name from `species` into `species2` for all situations where none of the conditions above it are met (e.g., if `species` is "Yellow Perch" then `species2` will be "Yellow Perch").

```{r}
PSDWRtest <- PSDWRtest |>
  mutate(species2=case_when(
           species=="Bluegill Sunfish" ~ "Bluegill",
           species=="Lean Lake Trout" ~ "Lake Trout",
           species=="Brown Trout" & location=="Trout Lake" ~  "Brown Trout (lotic)",
           species=="Brown Trout" & location=="Brushy Creek" ~  "Brown Trout (lentic)",
           species=="Brook Trout" & location=="Trout Lake" ~  "Brook Trout (lotic)",
           species=="Muskellunge" & sex=="M" ~ "Muskellunge (male)",
           species=="Muskellunge" & sex=="F" ~ "Muskellunge (female)",
           species=="Muskellunge" & sex=="U" ~ "Muskellunge (overall)",
           species=="Muskellunge" & is.na(sex) ~ "Muskellunge (overall)",
           species=="Walleye" & len>=150 ~ "Walleye (overall)",
           species=="Walleye" & len<150 ~ "Walleye (30-149 mm)",
           .default=species
         ))
peek(PSDWRtest,n=20)
```

The relative weights for each fish (for species with standard weight equations) is added to this data.drame with `wrAdd()` as described above, noting the use of the "new" species variable `species2`. Further note the use of `wsOpts=` to define which reference quantile to use for the Ruffe standard weight equation.

```{r}
PSDWRtest$wr <- wrAdd(wt~len+species2,data=PSDWRtest,
                      WsOpts=list(Ruffe=list(ref=75)))
peek(PSDWRtest,n=20)
```


# Summarizing Relative Weights
## Single Species
As an illustration, the mean and standard deviation of relative weights for Cisco (in `CiscoTL` from above) are computed below (along with `validn` which is the number of non-`NA` (i.e., non-missing) relative weights).^[More interesting summaries and plots may be constructed with these data. This is just an example of summarizing relative weights for a sample.] These results suggest that this population of Cisco is substantially "skinnier" (i.e., under-weight) than the standard for the species (because the mean relative weight is substantially less than 100).

```{r}
CiscoTL |>
  summarize(validn=sum(!is.na(Wr)),
            mnWr=mean(Wr,na.rm=TRUE),
            sdWr=sd(Wr,na.rm=TRUE))
```

## Multiple Species
The mean and sd of relative weight for all species in `PSDWRtest` from above is shown below. It is important to carefully examine these results because a species could have no mean relative weight calculated because there is no standard weight equation for that species (in `WSlit`) or because the species name is not spelled as expected (in `WSlit`). The only species below with no mean relative weight (i.e., Iowa Darter) does, indeed, not have a known standard weight equation.

```{r}
PSDWRtest |>
  group_by(species,location) |>
  summarize(validn=sum(!is.na(wr)),
            mnWr=mean(wr,na.rm=TRUE),
            sdWr=sd(wr,na.rm=TRUE)) |>
  as.data.frame()
```
