---
title: "Computing PSD and Relative Weight Metrics in FSA"
author: "Derek H. Ogle"
date: "`r Sys.Date()`"
format: 
  html: 
    toc: true
    toc-depth: 2
    reference-location: margin
    fig-cap-location: bottom
    tbl-cap-location: top
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

# Introduction
Other vignettes introduced using `FSA` to compute [proportional size distribution (PSD)](https://fishr-core-team.github.io/FSA/articles/Computing_PSDs.html) and [relative weight (i.e., condition)](https://fishr-core-team.github.io/FSA/articles/Computing_Relative_Weights.html) metrics. Both of those vignettes mentioned some peculiarities in their methodology related to when PSDs and relative weights were computed concurrently on the same data that contained some "issues." These peculiarities and issues are explained further and how to handle them are illustrated here. I urge you to explore those other vignettes before continuing here.

The following packages are used herein. Note that the `FSA` functions described here were modified after version 0.9.6 and are thus **specific to FSA >v0.9.6**.

```{r}
#| label: setup
#| results: hide
#| message: false
library(FSA)
library(dplyr)  # mutate, select, filter, case_when
```

# Data
The hypothetical `PSDWRtest` data.frame distributed with `FSA` was created to purposely provide data for a variety of species that were realistic but created some issues for the `psdAdd()` and `wrAdd()` functions described in the other vignettes.

```{r}
peek(PSDWRtest,n=20)
```

Ultimately it is best that you have a full understanding of the issues that may arise with your data by carefully examining your data and understanding the Gabelhouse (GH) length categories and relative weight equations for the species in your data. There are several "issues" that need to be addressed with the `PSDWRtest` data.

First, Bluegill and Lake Trout exist in `PSDlit` and `WSlit` but appear as "Bluegill Sunfish" and "Lean Lake Trout" in `PSDWRtest`.

Second, Brook Trout were sampled from a lotic ("Trout Lake") system and Brown Trout were sampled from a lotic ("Trout Lake") and a lentic ("Brush Creek") system. There are separate GH length categories and standard weight equations for these sub-groups for each species. 

Third, Muskellunge and Walleye provide a challenge because standard weight equations differ among sub-groups for each of these species, but the GH length categories do not differ among sub-groups. Walleye use separate standard weight equations depending on whether the individual's total length is less than 150 mm or not. Muskellunge have a sex-specific standard weight equation for when sex is known, but an overall equation for when sex is unknown. As will be seen, a new species variable will be created that incorporates the subgroup name into the "species name" for calculating standard weights (as was illustrated in [this vignette](https://fishr-core-team.github.io/FSA/articles/Computing_Relative_Weights.html)). Because the GH length categories don't differ by sub-group for either of these species, these new names are not needed when finding the GH length categories. However, for simplicity (as illustrated further below), these GH length categories have been duplicated for the sub-groups of each species and labelled with the combined species and sub-group name. This is illustrated below for muskellunge.

```{r}
wsVal("Muskellunge (overall)")  # different from next two
wsVal("Muskellunge (female)")
wsVal("Muskellunge (male)")
psdVal("Muskellunge (overall)") # exact same as next two
psdVal("Muskellunge (female)")
psdVal("Muskellunge (male)")
```

Fourth, Ruffe for which the reference quantile of the standard weight equation must be specified.

# Adding GH Length Category and Wr Variables
The easiest way to deal with all of these "issues" except for the one related to Ruffe is to create a new "species" variable (i.e., `species2` below) that appends the specific sub-groups in parentheses to the species name. There are a variety of ways to do this and which way (is best or works) may depend on the specifics of the situation. Here, we use `case_when()` from `dplyr` with a series of statements that begin with a "condition" to the left of the `~` and new species "name" for that condition to the right of the `~`. The `.default=species` at the end puts the name from `species` into `species2` for all situations where none of the conditions above it are met (e.g., if `species` is "Yellow Perch" then `species2` will be "Yellow Perch").

```{r}
PSDWRtest <- PSDWRtest |>
  mutate(species2=case_when(
           species=="Bluegill Sunfish" ~ "Bluegill",
           species=="Lean Lake Trout" ~ "Lake Trout",
           species=="Brown Trout" & location=="Trout Lake" ~  "Brown Trout (lotic)",
           species=="Brown Trout" & location=="Brushy Creek" ~  "Brown Trout (lentic)",
           species=="Brook Trout" & location=="Trout Lake" ~  "Brook Trout (lotic)",
           species=="Muskellunge" & sex=="M" ~ "Muskellunge (male)",
           species=="Muskellunge" & sex=="F" ~ "Muskellunge (female)",
           species=="Muskellunge" & sex=="U" ~ "Muskellunge (overall)",
           species=="Muskellunge" & is.na(sex) ~ "Muskellunge (overall)",
           species=="Walleye" & len>=150 ~ "Walleye (overall)",
           species=="Walleye" & len<150 ~ "Walleye (30-149 mm)",
           .default=species
         ))
peek(PSDWRtest,n=20)
```

The GH length categories and relative weights for each fish (for species with standard weight equations) is added to this data.drame with `psdAdd()` and `wrAdd()` as described in the other vignettes, specifically noting the use of the "new" species variable `species2`. Further note the use of `wsOpts=` to define which reference quantile to use for the Ruffe standard weight equation.

```{r}
PSDWRtest$psd <- psdAdd(len~species2,data=PSDWRtest)
PSDWRtest$wr <- wrAdd(wt~len+species2,data=PSDWRtest,
                      WsOpts=list(Ruffe=list(ref=75)))
peek(PSDWRtest,n=20)
```

